---
title: "Get started (R)"
---

## Installation

Install from [r-universe](https://bgreenwell.r-universe.dev/) (once the
build finishes syncing after registration):

```r
install.packages("statlingo", repos = c("https://bgreenwell.r-universe.dev", "https://cloud.r-project.org"))
```

Or install the development version directly from GitHub:

```r
# install.packages("pak")
pak::pak("bgreenwell/statlingo/r")
```

You'll also need the [ellmer](https://cran.r-project.org/package=ellmer)
package to create an LLM client:

```r
install.packages("ellmer")
```

## Quick example

First, load the package, fit a linear model, and define the context of your data:

```{r}
#| eval: false
library(statlingo)

# Fit a simple linear model
fm <- lm(dist ~ speed, data = cars)

context <- "
The data give the speed of cars (mph) and the distances taken to stop
(ft). Note that the data were recorded in the 1920s!
"
```

This fits a basic linear regression of stopping distance on speed using the built-in `cars` dataset and captures the research context.

Next, initialize the `ellmer` chat client using Google Gemini:

```{r}
#| eval: false
# Create an ellmer chat client using Google Gemini
client <- ellmer::chat_google_gemini()
```

This creates a chat client instance configured to use Gemini models. The client is passed to `explain()` without mutating the original client state.

Generate the natural language explanation using `explain()`, targeting a student audience:

```{r}
#| eval: false
# Get an explanation tailored for a student audience
ex <- explain(
  fm,
  client = client,
  context = context,
  audience = "student",
  verbosity = "detailed"
)

print(ex)
```

The returned explanation object prints out a structured, student-oriented summary of the regression model, detailing coefficients, overall fit, and checkable assumptions.

Finally, request diagnostic code recommendations based on the generated explanation using `suggest_code()`:

```{r}
#| eval: false
# Suggest next diagnostic code steps based on the explanation
suggest_code(ex)
```

This prints recommended R code snippets to check residual diagnostics, multicollinearity, autocorrelation, and constant variance.

## Learn more

* [Function reference](r/reference/index.html)
* [Source code](https://github.com/bgreenwell/statlingo/tree/main/r)
