---
title: "Local LLM Support"
---

Running `statlingo` with local, offline Large Language Models (LLMs) is an excellent way to avoid internet dependencies, external API costs, and API key requirements. This is particularly useful in classroom settings, private enterprise environments, or when working with sensitive data.

This guide shows how to run `statlingo` offline using locally hosted models via **Ollama** or **Llama.cpp**.

---

## 1. Using Ollama (Recommended)

[Ollama](https://ollama.com/) is a lightweight, easy-to-use framework for running LLMs locally on macOS, Linux, and Windows.

### Step A: Set up Ollama and Pull a Model
1. Download and install Ollama from [ollama.com](https://ollama.com/).
2. Start the Ollama application.
3. Open your terminal and pull a small, lightweight model of your choice:
   ```bash
   # Llama 3.2 (1-billion parameters) - Recommended for speed and basic capability
   ollama pull llama3.2:1b
   
   # Gemma 2 (2-billion parameters) - Recommended for higher quality explanations
   ollama pull gemma2:2b
   ```

### Step B: Run with the R Package
The R package integrates with Ollama via `ellmer::chat_ollama()`.

```{r}
#| eval: false
library(statlingo)
library(ellmer)

# 1. Fit your statistical model
model <- lm(dist ~ speed, data = cars)

# 2. Initialize the local Ollama client (default port is 11434)
client <- chat_ollama(model = "llama3.2:1b")

# 3. Generate the natural language explanation
explanation <- explain(model, client = client, audience = "student")
print(explanation)
```

### Step C: Run with the Python Package
The Python package integrates with Ollama via `chatlas.ChatOllama`.

```{python}
#| eval: false
import statsmodels.api as sm
from chatlas import ChatOllama
from statlingo import explain

# 1. Fit your statistical model
duncan_data = sm.datasets.get_rdataset("Duncan", "carData").data
y = duncan_data["prestige"]
X = sm.add_constant(duncan_data[["income"]])
model = sm.OLS(y, X).fit()

# 2. Initialize the local Ollama client (default port is 11434)
client = ChatOllama(model="llama3.2:1b")

# 3. Generate the natural language explanation
explanation = explain(model_object=model, client=client, audience="student")
print(explanation["text"])
```

---

## 2. Using Llama.cpp (via OpenAI-Compatible Server)

[Llama.cpp](https://github.com/ggerganov/llama.cpp) allows highly optimized local LLM inference. You can use it with `statlingo` by running its built-in HTTP server (`llama-server`), which exposes an API compatible with OpenAI's endpoint.

### Step A: Start the Llama.cpp Server
Run the local server on your machine, specifying your downloaded GGUF model file and a port (e.g. `8080`):
```bash
./llama-server -m models/llama-3.2-1b-instruct.Q4_K_M.gguf -c 2048 --port 8080
```

### Step B: Run with the R Package
Use `ellmer::chat_openai_compatible()` to point to your local Llama.cpp server port:

```{r}
#| eval: false
library(statlingo)
library(ellmer)

model <- lm(dist ~ speed, data = cars)

# Connect to the local llama-server
client <- chat_openai_compatible(
  model = "local-model",
  base_url = "http://localhost:8080/v1",
  api_key = "not-needed"
)

explanation <- explain(model, client = client, audience = "student")
print(explanation)
```

### Step C: Run with the Python Package
Use `chatlas.ChatOpenAICompletions` to point to your local Llama.cpp server port:

```{python}
#| eval: false
import statsmodels.api as sm
from chatlas import ChatOpenAICompletions
from statlingo import explain

# Fit a model
duncan_data = sm.datasets.get_rdataset("Duncan", "carData").data
y = duncan_data["prestige"]
X = sm.add_constant(duncan_data[["income"]])
model = sm.OLS(y, X).fit()

# Connect to the local llama-server
client = ChatOpenAICompletions(
    model="local-model",
    base_url="http://localhost:8080/v1",
    api_key="not-needed"
)

explanation = explain(model_object=model, client=client, audience="student")
print(explanation["text"])
```
