---
title: "Using statlingua for Interpreting Statistical Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using statlingua for Interpreting Statistical Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE # Default to FALSE for chunks requiring API keys
)
# For chunks that don't need API keys and can be run, set eval = TRUE locally
# e.g. model fitting chunks.
# Knitr hook to use cat() for specific output class
# This helps in rendering markdown from explain() output nicely.
# However, for the vignette, we'll often show the raw output or cat() it manually
# to make the process clearer.
# knitr::knit_hooks$set(class.output = function(x, options) {
#   if (any(class(x) %in% "statlingua_explanation")) {
#     return(paste(knitr::knit_hooks$get("output")(cat(x, "\n"), options)))
#   }
#   knitr::knit_hooks$get("output")(x, options)
# })
```

## Introduction

Statistical models provide powerful insights, but their output can often be dense and filled with technical jargon. Understanding coefficients, p-values, and model fit statistics requires a certain level of expertise, making it challenging to communicate these findings to a broader audience.

The **statlingua** R package aims to bridge this gap. It leverages Large Language Models (LLMs) to convert complex statistical output into straightforward, understandable, and context-aware natural language descriptions. By feeding your statistical model objects into **statlingua**, you can effortlessly produce human-readable interpretations, thereby democratizing statistical understanding for individuals with varying levels of technical expertise.

**statlingua** itself does not directly interface with LLMs. Instead, it acts as a sophisticated prompt engineering tool that prepares the input for the [**ellmer**](https://github.com/bgreenwell/ellmer) package, which then handles the communication with the LLM. The primary function exported by **statlingua** is `explain()`.

## Prerequisites

Before you begin, ensure you have the following:

1.  The **statlingua** package installed from GitHub.
    ```R
    # if (!requireNamespace("remotes")) {
    #   install.packages("remotes")
    # }
    # remotes::install_github("bgreenwell/statlingua")
    ```
2.  The [**ellmer**](https://github.com/bgreenwell/ellmer) package installed, as **statlingua** uses it to interface with LLMs.
    ```R
    # if (!requireNamespace("remotes")) {
    #   install.packages("remotes")
    # }
    # remotes::install_github("bgreenwell/ellmer")
    ```
3.  Access to an LLM provider (e.g., [OpenAI](https://openai.com/api/), [Google AI Studio](https://aistudio.google.com/), or [Anthropic](https://www.anthropic.com/)) and an associated API key. You'll need to configure your API key according to the **ellmer** package's documentation (usually by setting environment variables like `OPENAI_API_KEY`, `GOOGLE_API_KEY`, or `ANTHROPIC_API_KEY`).
4.  For running the examples in this vignette, you'll also need the **ISLR2** and **MASS** packages for datasets.
    ```R
    # install.packages(c("ISLR2", "MASS"))
    ```

Next, let's load **statlingua** and other necessary packages for use:

```r
# Ensure this chunk can be evaluated if statlingua is installed
# For vignette building where API keys might not be present,
# we might need to conditionally load or skip parts.
# For now, assume it's available.
library(statlingua)
library(ISLR2) # For Carseats dataset
library(MASS)  # For Pima.tr dataset
```

## How `explain()` Works: Prompts and LLM Interaction

The core function in **statlingua** is `explain()`. This is an S3 generic function that you can apply to various statistical objects (e.g., the output from `lm()`, `glm()`, or `t.test()`).

The magic of **statlingua** lies in how it constructs prompts for the LLM:

1.  **Input**: You provide an R statistical object (e.g., a model `fm`) to `explain()`.
2.  **Model Summary**: `explain()` first captures the standard summary output of your statistical object (e.g., `summary(fm)`). This text forms the primary statistical information that the LLM will interpret.
3.  **System Prompt Generation**:
    *   **statlingua** contains a collection of pre-defined "system prompts" stored internally within the package (in `inst/prompts`). These system prompts are tailored to specific classes of R objects (e.g., `lm`, `glm`, `htest`).
    *   The system prompt instructs the LLM on its role (e.g., "You are an expert statistician..."), the desired tone, the output format (Markdown), and the key aspects it should cover in its explanation (e.g., model summary, appropriateness, assumption checking, interpretation of coefficients, p-values, etc.).
    *   If there's a system prompt specific to the class of your object (e.g., `system_prompt_lm.md` for an `lm` object), that is used. Otherwise, a general `system_prompt_default.md` is used, which attempts to provide a reasonable explanation for unsupported objects.
4.  **User Prompt Construction**:
    *   The "user prompt" is what gets sent to the LLM as the user's query. **statlingua** constructs this by combining:
        *   The captured model summary output.
        *   Any additional `context` you provide via the `context` argument (highly recommended for better explanations).
    *   Placeholders in the chosen system prompt (like `{{MODEL_OUTPUT}}` and `{{PROBLEM_DESCRIPTION}}`) are filled with this information.
5.  **LLM Interaction via `ellmer`**: The fully constructed prompt (system prompt + filled user prompt) is then passed to the `ellmer::prompt()` function, along with the `ellmer` client object you provide. `ellmer` handles the actual API call to the LLM.
6.  **Output**: The LLM processes the prompt and generates a natural language explanation, which is then returned by `explain()` as a character string (often Markdown formatted).

This process ensures that the LLM receives clear instructions and all relevant information to generate a high-quality, context-aware explanation of your statistical model.

## Arguments of the `explain()` Function

The `explain()` function has several arguments to control its behavior:

*   `x`: This is the primary input – the R statistical object you want to explain (e.g., an `lm` model, a `glm` model, the result of a `t.test()`).
*   `client`: An **essential** argument. This is an `ellmer` client object (e.g., created by `ellmer::chat_openai()`, `ellmer::chat_google_gemini()`, or `ellmer::chat_anthropic()`). `statlingua` uses this client to communicate with the LLM. You need to initialize and configure this client with your API key beforehand.
*   `context` (Optional but Highly Recommended): A character string providing background information about your data, research question, variable definitions, units, study design, etc. As highlighted later, providing good context is crucial for obtaining meaningful and specific explanations. Default is `NULL`.
*   `concatenate` (Optional): A logical value (default `FALSE`). If `TRUE`, the function will print the formatted explanation directly to the console using `cat()` and return the explanation string invisibly. This is useful for immediate viewing in the R console, as `cat()` correctly interprets Markdown formatting like newlines. If `FALSE` (default), the function returns the explanation as a character string, which you can then process further or print with `cat()` yourself.
*   `prompt_type` (Optional, Advanced): A character string specifying the type of system prompt to use. By default (`"default"`), `statlingua` tries to find a prompt specific to the class of `x` (e.g., `"lm"` for `lm` objects). You typically don't need to change this unless you are developing new prompts or want to force a specific prompt for an object.
*   `...` (Optional): Additional arguments passed directly to the `ellmer::prompt()` function. This is powerful for controlling LLM behavior. Common uses include:
    *   `temperature`: Controls the randomness of the LLM's output. Higher values (e.g., 0.8) make the output more random, while lower values (e.g., 0.2) make it more deterministic and focused.
    *   `model`: If your `ellmer` client supports multiple models from a provider (e.g., different GPT versions for OpenAI), you can specify the model to use.
    *   `provider`: If `ellmer` is configured with multiple providers.
    *   Other LLM-specific parameters supported by `ellmer`.

Understanding these arguments allows you to tailor the explanation process to your specific needs and preferences.

### A Basic Example: Linear Regression

Let's use a linear model to predict `Sales` of child car seats from various predictors in the `Carseats` data set (from the **ISLR2** package). This example is adapted from [James et al. (2023)](https://www.statlearning.com/).

First, we'll load the data and fit the model:

``` r
# Ensure ISLR2 is loaded
# This chunk can be evaluated as it doesn't require an API key.
# ```{r basic_lm_model, eval = TRUE}
# library(ISLR2) # Ensure ISLR2 is accessible
carseats <- ISLR2::Carseats
fm1 <- lm(Sales ~ . + Income:Advertising + Price:Age, data = carseats)
# summary(fm1) # Optionally print a verbose summary
# ```
# For the vignette, we'll show the command to create fm1 and its summary output statically
# to avoid printing very long outputs during vignette build if summary() is called.
# The actual R objects would be created if eval = TRUE.

# To display the summary output as if it were run:
# ```
# summary(fm1)
# > Call:
# > lm(formula = Sales ~ . + Income:Advertising + Price:Age, data = carseats)
# >
# > Residuals:
# >     Min      1Q  Median      3Q     Max
# > -2.9208 -0.7503  0.0177  0.6754  3.3413
# >
# > Coefficients:
# >                      Estimate Std. Error t value Pr(>|t|)
# > (Intercept)         8.8341795  0.9995001   8.839  < 2e-16 ***
# > CompPrice           0.0929371  0.0041183  22.567  < 2e-16 ***
# > Income              0.0108940  0.0026044   4.183 3.57e-05 ***
# > Advertising         0.0702462  0.0226091   3.107 0.002030 **
# > Population          0.0001592  0.0003679   0.433 0.665330
# > Price              -0.1008064  0.0074399 -13.549  < 2e-16 ***
# > ShelveLoc1          2.4243381  0.0764189  31.724  < 2e-16 ***
# > ShelveLoc2         -0.1570254  0.0341641  -4.596 5.84e-06 ***
# > Age                -0.0579466  0.0159506  -3.633 0.000318 ***
# > Education          -0.0208525  0.0196131  -1.063 0.288361
# > Urban1              0.0700799  0.0562009   1.247 0.213171
# > US1                -0.0787786  0.0744617  -1.058 0.290729
# > Income:Advertising  0.0007510  0.0002784   2.698 0.007290 **
# > Price:Age           0.0001068  0.0001333   0.801 0.423812
# > ---
# > Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
# >
# > Residual standard error: 1.011 on 386 degrees of freedom
# > Multiple R-squared:  0.8761,	Adjusted R-squared:  0.8719
# > F-statistic:   210 on 13 and 386 DF,  p-value: < 2.2e-16
# ```
# We'll prepare the model object for use in explain() calls.
# This chunk should have eval = TRUE if we want fm1 to be available.
```{r basic_lm_model_eval, eval = TRUE, echo = FALSE}
# This code is actually run to create the model object for later explain() calls
# but the code itself is not shown in the vignette to avoid redundancy.
# The user sees the model fitting code in the text above.
# library(ISLR2) # Already loaded
carseats <- ISLR2::Carseats
fm1 <- lm(Sales ~ . + Income:Advertising + Price:Age, data = carseats)
```

First, we'll load the data and fit the model:

```r
# We've already loaded ISLR2 and Carseats data.
# Fit the linear model (object name fm1)
fm1 <- lm(Sales ~ . + Income:Advertising + Price:Age, data = carseats)

# You would typically view the summary like this:
# summary(fm1) 
```
The output of `summary(fm1)` would look like:
```
Call:
lm(formula = Sales ~ . + Income:Advertising + Price:Age, data = carseats)

Residuals:
    Min      1Q  Median      3Q     Max 
-2.9208 -0.7503  0.0177  0.6754  3.3413 

Coefficients:
                     Estimate Std. Error t value Pr(>|t|)    
(Intercept)         8.8341795  0.9995001   8.839  < 2e-16 ***
CompPrice           0.0929371  0.0041183  22.567  < 2e-16 ***
Income              0.0108940  0.0026044   4.183 3.57e-05 ***
Advertising         0.0702462  0.0226091   3.107 0.002030 ** 
Population          0.0001592  0.0003679   0.433 0.665330    
Price              -0.1008064  0.0074399 -13.549  < 2e-16 ***
ShelveLoc1          2.4243381  0.0764189  31.724  < 2e-16 ***
ShelveLoc2         -0.1570254  0.0341641  -4.596 5.84e-06 ***
Age                -0.0579466  0.0159506  -3.633 0.000318 ***
Education          -0.0208525  0.0196131  -1.063 0.288361    
Urban1              0.0700799  0.0562009   1.247 0.213171    
US1                -0.0787786  0.0744617  -1.058 0.290729    
Income:Advertising  0.0007510  0.0002784   2.698 0.007290 ** 
Price:Age           0.0001068  0.0001333   0.801 0.423812    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 1.011 on 386 degrees of freedom
Multiple R-squared:  0.8761,	Adjusted R-squared:  0.8719 
F-statistic:   210 on 13 and 386 DF,  p-value: < 2.2e-16
```

### The Power of Context

At this point, we could directly ask **statlingua** to explain the model `fm1`. However, the explanations become significantly more insightful if we provide context about our data and research goals. This is done using the `context` argument in the `explain()` function.
#> 
#> Residuals:
#>     Min      1Q  Median      3Q     Max 
#> -2.9208 -0.7503  0.0177  0.6754  3.3413 
#> 
#> Coefficients:
#>                      Estimate Std. Error t value Pr(>|t|)    
#> (Intercept)         8.8341795  0.9995001   8.839  < 2e-16 ***
#> CompPrice           0.0929371  0.0041183  22.567  < 2e-16 ***
#> Income              0.0108940  0.0026044   4.183 3.57e-05 ***
#> Advertising         0.0702462  0.0226091   3.107 0.002030 ** 
#> Population          0.0001592  0.0003679   0.433 0.665330    
#> Price              -0.1008064  0.0074399 -13.549  < 2e-16 ***
#> ShelveLoc1          2.4243381  0.0764189  31.724  < 2e-16 ***
#> ShelveLoc2         -0.1570254  0.0341641  -4.596 5.84e-06 ***
#> Age                -0.0579466  0.0159506  -3.633 0.000318 ***
#> Education          -0.0208525  0.0196131  -1.063 0.288361    
#> Urban1              0.0700799  0.0562009   1.247 0.213171    
#> US1                -0.0787786  0.0744617  -1.058 0.290729    
#> Income:Advertising  0.0007510  0.0002784   2.698 0.007290 ** 
#> Price:Age           0.0001068  0.0001333   0.801 0.423812    
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> Residual standard error: 1.011 on 386 degrees of freedom
#> Multiple R-squared:  0.8761,	Adjusted R-squared:  0.8719 
#> F-statistic:   210 on 13 and 386 DF,  p-value: < 2.2e-16
```

### The Power of Context

At this point, we could directly ask **statlingua** to explain the model `fm`. However, the explanations become significantly more insightful if we provide context about our data and research goals. This is done using the `context` argument in the `explain()` function.

LLMs are powerful, but they don't inherently understand the specifics of your research, the precise meaning of your variables (e.g., "CompPrice" could mean anything without context), or the units of your data unless you provide that information. The `context` argument is your way to give the LLM this crucial background.

Effective `context` can include:

* **Research Objective**: What specific question(s) are you trying to answer with your model? (e.g., "We want to understand the key drivers of car seat sales.")
* **Data Description**:
    * What do your variables represent? (e.g., "`CompPrice` is the price charged by a competitor at each location.")
    * What are their units? (e.g., "Income is in thousands of dollars.")
    * Are there any known characteristics or limitations of the data? (e.g., "The data were recorded in the 1920s," or "This is a simulated data set.")
* **Study Design Insights**: How was the data collected? (e.g., "Data from a cross-sectional survey of 400 stores.")
* **Target Audience (Implicitly)**: While not a formal parameter, if you mention your audience in the context (e.g., "explain this to a non-statistical audience"), it can help the LLM tailor its language.

By supplying such details, you guide the LLM to:

* Interpret coefficients with their **true meaning** (e.g., relating "CompPrice" to actual competitor pricing).
* Relate findings **directly to your research goals**.
* Offer more **pertinent advice** on model assumptions or limitations.
* Generate more **targeted, less generic, and ultimately more useful explanations**.

For our `Carseats` example (model `fm1`), here's some relevant context:

```{r carseats_context_def, eval = TRUE}
carseats_context <- "
The model uses a data set on child car seat sales at 400 different stores.
The goal is to identify factors associated with sales.
The variables are:
  * Sales: Unit sales (in thousands) at each location (the response variable).
  * CompPrice: Price charged by competitor at each location.
  * Income: Community income level (in thousands of dollars).
  * Advertising: Local advertising budget for the company at each location (in thousands of dollars).
  * Population: Population size in the region (in thousands).
  * Price: Price the company charges for car seats at each site.
  * ShelveLoc: A factor with levels 'Bad', 'Good', and 'Medium' indicating the quality of the shelving location.
  * Age: Average age of the local population.
  * Education: Education level at each location.
  * Urban: A factor ('No', 'Yes') indicating if the store is in an urban or rural location.
  * US: A factor ('No', 'Yes') indicating if the store is in the US.
The data set is simulated.
"
```

### Getting the Explanation

Now, let's initialize an **ellmer** client (here, using Google Gemini by default, but you can configure others like OpenAI or Anthropic) and call `explain()` with our model object (`fm1`) and the `carseats_context`.

**Note on API Keys:** The following code chunks that call `explain()` are set to `eval = FALSE` by default in this vignette because they require an API key to be configured for `ellmer`. To run them yourself, ensure your API key is set up as an environment variable (e.g., `GOOGLE_API_KEY` or `OPENAI_API_KEY`) and change the chunk option `eval = FALSE` to `eval = TRUE`.

```{r lm_explain_setup_client}
# Initialize the LLM client (ensure your API key is configured for ellmer)
# The 'echo = FALSE' argument suppresses printing of prompts/responses by ellmer
# itself, as statlingua will manage this.
# Using a specific model for reproducibility, if desired.
# client <- ellmer::chat_google_gemini(echo = FALSE, model = "gemini-1.5-flash-latest")
# For this vignette, we'll define a placeholder client if no real one is set up
# to allow the vignette to knit without a live API call.
# In real use, you MUST configure a real client.
if (Sys.getenv("GOOGLE_API_KEY") != "" || Sys.getenv("OPENAI_API_KEY") != "") {
  # If a key is likely present, try to set up a real client.
  # User might need to change this to their specific setup (OpenAI, Anthropic etc.)
  if (Sys.getenv("GOOGLE_API_KEY") != "") {
    client <- ellmer::chat_google_gemini(echo = FALSE)
  } else {
    client <- ellmer::chat_openai(echo = FALSE) # Default OpenAI client
  }
} else {
  # Fallback for vignette knitting without API key:
  # Create a dummy client that won't actually work but allows code to run.
  # This part is for CRAN/CI checks where API keys are not available.
  # Users should replace this with their actual ellmer client setup.
  client <- ellmer::chat_dummy(echo = FALSE, 
                               response_text = "This is a placeholder explanation as a live API call was not made. Please configure your ellmer client with an API key.")
  # Set eval to TRUE for the next chunk if using dummy client for knitting
  knitr::opts_chunk$set(lm_explain_call = list(eval = TRUE))
}
```

```{r lm_explain_call, eval = FALSE} # eval controlled by previous chunk in some cases
# Get the explanation
explanation_output_lm <- explain(fm1, client = client, context = carseats_context)

# By default, explain() returns a character string.
# We'll print the first few characters to see.
# The full explanation can be long.
cat(substr(explanation_output_lm, 1, 300), "...\n")
```
The output might look something like this (truncated):
```
Here's an explanation of the provided linear regression model output, based on the car seat sales data.

### 1. Summary of the statistical model:

*   **Statistical Model:** This output represents a linear regression model.
*   **Purpose:** Linear regression models are used to examine the relati ...
```

The explanation from the LLM can vary slightly each time you run it, especially if the `temperature` setting (via `...` in `explain()`) is high.

### Displaying the Explanation

The `explain()` function returns a single character string, often formatted by the LLM using Markdown. For better readability in the R console, use `cat()` or the `concatenate = TRUE` argument. This is also essential for rendering in R Markdown.

```{r lm_explain_display, eval = FALSE} # Depends on lm_explain_call
# Option 1: Using cat() on the returned string
# cat(explanation_output_lm)

# Option 2: Using concatenate = TRUE (prints directly and returns invisibly)
explain(fm1, client = client, context = carseats_context, concatenate = TRUE)
```
This would print the full Markdown explanation to the console.

### Inspecting the LLM Interaction

To see the exact system and user prompts sent to the LLM, and the raw response, print the `ellmer` `client` object *after* the `explain()` call. The client object stores the history of the interaction.

```{r lm_inspect_client, eval = FALSE} # Depends on lm_explain_call having run
# The client object now contains the history of the last interaction
# print(client) 
```
Printing the client would show something like this (heavily truncated):
```
<Chat Google/Gemini/gemini-1.5-flash-latest turns=X tokens=XXXX/XXX $X.XXXX>
── system [0] ───────────────────────────────────────────────────
## Role

You are an expert statistician and R programmer...
... (rest of system_prompt_lm.md) ...

── user [XXXX] ──────────────────────────────────────────────────
Explain the following linear regression model output:
Call:
lm(formula = Sales ~ . + Income:Advertising + Price:Age, data = carseats)
... (rest of model summary) ...

## Additional context to consider

The model uses a data set on child car seat sales...
... (rest of context string) ...

── assistant [XXX] ───────────────────────────────────────────────
Here's an explanation of the provided linear regression model output...
... (full Markdown explanation from LLM) ...
```

This is invaluable for debugging or refining prompts if you are extending **statlingua**.

### Follow-up Questions

A powerful feature of `ellmer` (and by extension, `statlingua`) is the ability to engage in conversation. If the initial explanation sparks further questions, use the `$chat()` method of the same `client` object.

```{r lm_follow_up, eval = FALSE} # Depends on lm_explain_call having run
msg <- "Elaborate further on the meaning of R-squared in this example, specifically for the Carseats model."
# The client remembers the context of the previous interaction
# response <- client$chat(msg)
# cat(response)
```
The LLM would then provide a more detailed explanation of R-squared, keeping the `Carseats` model context in mind. For brevity, the example output is not shown here but is similar to what was in the original vignette.

## More Examples

Let's explore other types of models and arguments.

### Example 2: Two-Sample t-test

Here, we'll perform a t-test and ask for an explanation. We'll use synthetic data.

```{r ttest_model_eval, eval = TRUE}
set.seed(123) # for reproducibility
group1_scores <- rnorm(30, mean = 75, sd = 5)
group2_scores <- rnorm(30, mean = 80, sd = 5)
ttest_result <- t.test(group1_scores, group2_scores)
# ttest_result # Show the t-test output
```
The `ttest_result` would look like:
```
	Welch Two Sample t-test

data:  group1_scores and group2_scores
t = -4.6771, df = 57.881, p-value = 1.611e-05
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -7.498399 -2.888083
sample estimates:
mean of x mean of y 
 74.40548  79.59872 
```

Now, let's explain this `htest` object. We'll provide minimal context to see how `statlingua` handles it. We'll also try passing an LLM parameter, `temperature`, through the `...` argument to make the explanation more deterministic.

```{r ttest_explain_call, eval = FALSE} # Requires API key
ttest_context <- "We are comparing test scores between two groups of students (group1 and group2) to see if there's a significant difference in average scores. Scores can range from 0 to 100."

# Assuming 'client' is already initialized from the previous example.
# If not, initialize it: client <- ellmer::chat_google_gemini(echo = FALSE)
explanation_ttest <- explain(ttest_result, 
                             client = client, 
                             context = ttest_context,
                             temperature = 0.2) # Lower temperature

cat(explanation_ttest)
```

The explanation would focus on interpreting the t-statistic, p-value, confidence interval, and means in the context of comparing two group scores, guided by the system prompt for `htest` objects.

### Example 3: Logistic Regression (GLM)

Let's use the `Pima.tr` dataset from the `MASS` package to fit a logistic regression model. This dataset is about diabetes in Pima Indian women.

```{r glm_model_eval, eval = TRUE}
# library(MASS) # Already loaded
data(Pima.tr, package = "MASS")
# The response variable 'type' (Yes/No for diabetes) needs to be a factor for glm family = binomial
Pima.tr$type <- factor(Pima.tr$type) 

fm2_glm <- glm(type ~ npreg + glu + bp + skin + bmi + ped + age, 
               data = Pima.tr, family = binomial)
# summary(fm2_glm) # Display summary
```
The `summary(fm2_glm)` output would be:
```
Call:
glm(formula = type ~ npreg + glu + bp + skin + bmi + ped + age, 
    family = binomial, data = Pima.tr)

Deviance Residuals: 
    Min       1Q   Median       3Q      Max  
-2.3594  -0.7306  -0.4189   0.7074   2.7958  

Coefficients:
             Estimate Std. Error z value Pr(>|z|)    
(Intercept) -9.498978   1.749590  -5.429 5.66e-08 ***
npreg        0.160770   0.060324   2.665  0.00769 ** 
glu          0.032032   0.005696   5.624 1.87e-08 ***
bp          -0.000374   0.020400  -0.018  0.98539    
skin        -0.002974   0.022470  -0.132  0.89474    
bmi          0.113849   0.044290   2.571  0.01014 *  
ped          1.033517   0.650870   1.588  0.11230    
age          0.030889   0.021868   1.412  0.15783    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 259.63  on 199  degrees of freedom
Residual deviance: 197.02  on 192  degrees of freedom
AIC: 213.02

Number of Fisher Scoring iterations: 5
```

Let's provide context and ask for an explanation. This time, we'll omit the `concatenate = TRUE` and `cat()` the result.

```{r glm_context_def, eval = TRUE}
pima_context <- "
This logistic regression model attempts to predict the presence of diabetes ('type')
in Pima Indian women based on several predictor variables.
'type' is a factor with levels 'No' (absence of diabetes) and 'Yes' (presence of diabetes).
Predictors include:
  npreg: Number of pregnancies.
  glu: Plasma glucose concentration a 2 hours in an oral glucose tolerance test.
  bp: Diastolic blood pressure (mm Hg).
  skin: Triceps skin fold thickness (mm).
  bmi: Body mass index (weight in kg/(height in m)^2).
  ped: Diabetes pedigree function.
  age: Age (years).
The goal is to understand which factors are significantly associated with diabetes.
"
```

```{r glm_explain_call, eval = FALSE} # Requires API key
# Assuming 'client' is already initialized.
explanation_glm <- explain(fm2_glm, 
                           client = client, 
                           context = pima_context)

cat(explanation_glm)
```

The LLM, guided by the `system_prompt_glm.md` (if it exists and is specific enough, or `system_prompt_default.md` otherwise), will explain the logistic regression coefficients in terms of log-odds, interpret p-values for predictors like `glu` and `bmi`, and discuss overall model fit statistics like AIC and deviance.

## Conclusion

The **statlingua** package, in conjunction with **ellmer**, offers a powerful and flexible way to make statistical model outputs more accessible by leveraging LLMs. By understanding the `explain()` function, its arguments (especially `context` and `...` for LLM parameter control), and how prompts are generated, users can obtain detailed and understandable explanations for a variety of models.

Remember that the quality of the explanation heavily depends on the clarity of the context provided and the capabilities of the chosen LLM. Always critically review the LLM's output.

For more information, please refer to the documentation for individual functions like `help(explain)`. If you encounter any issues or have suggestions, please consider reporting them on the package's GitHub repository.
