---
title: "examples"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**WARNING:** This article is incomplete and still very much a work in progress!

This article aims to show a wide variety of examples in which we use Large Language Models (LLMs) to help explain the output from various statistical tests and models.

## Explaining the output from a two-sample t-test

The following example is taken from a [tutorial on paired t-tests](https://www.jmp.com/en/statistics-knowledge-portal/t-test/paired-t-test). Here we use an independent two-sample t-test to (inappropriately) analyze paired data.

```{r ex-t-test, cache=TRUE, results="asis"}
library(statlingua)

# Define additional context to pass to `explain()`; this should include 
# any additional background information about the data and research question.
context <- "
An instructor wants to use two exams in her classes next year. This year, she gives both exams to the students. She wants to know if the exams are equally difficult and wants to check this by comparing the two sets of scores. Here is the data:

 student exam_1_score exam_2_score
     Bob           63           69
    Nina           65           65
     Tim           56           62
    Kate          100           91
  Alonzo           88           78
    Jose           83           87
  Nikhil           77           79
   Julia           92           88
   Tohru           90           85
 Michael           84           92
    Jean           68           69
   Indra           74           81
   Susan           87           84
   Allen           64           75
    Paul           71           84
  Edwina           88           82
"

# Create the data set
exam_scores <- tibble::tribble(
  ~student,  ~exam_1_score, ~exam_2_score,
  "Bob",     63,            69,
  "Nina",    65,            65,
  "Tim",     56,            62,
  "Kate",    100,           91,
  "Alonzo",  88,            78,
  "Jose",    83,            87,
  "Nikhil",  77,            79,
  "Julia",   92,            88,
  "Tohru",   90,            85,
  "Michael", 84,            92,
  "Jean",    68,            69,
  "Indra",   74,            81,
  "Susan",   87,            84,
  "Allen",   64,            75,
  "Paul",    71,            84,
  "Edwina",  88,            82
)

# Run a two-sample t-test
(tt <- t.test(exam_scores$exam_1_score, y = exam_scores$exam_2_score))

# Inititalize client
client <- ellmer::chat_google_gemini(echo = "none")
ex <- explain(tt, client = client, context = context)
cat(ex)
```
